\bf A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
Modify the blind option.

\documentclass[12pt]{article}
\usepackage{amsmath}
%\usepackage{longtable}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
\usepackage{float}
\usepackage{url} % not crucial - just used below for the URL
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

A robust linear regression model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002) is used to treat the extreme values in the data. The robustness weight is used to determine if a value should be replaced with the fitted value from the model. This is constructed for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively. An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016). However, when we tested this, the fit failed to converge. The full code for this is shown at \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html} but also described in detail next.

The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively. Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021). The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.

To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done. Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles. Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals. A threshold of 0.12 was chosen. That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called \texttt{is\_pred}, so that this change can be monitored in the downstream analyses.

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.

@@

\begin{itemize}
\item
  \textbf{Codes}: R script to reproduce data tidying and cleaning is available in this \href{https://github.com/numbats/yowie/articles/process-data.html}{page}.
\item
  \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed \href{https://github.com/numbats/yowie}{here}.
\item
  \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model. This app can be accessed \href{https://ebsmonash.shinyapps.io/yowie_app/}{here} with the source code provided \href{https://github.com/numbats/yowie/tree/master/app}{here}.
\end{itemize}

@@

\leavevmode\hypertarget{ref-rlm}{}%
UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
%\usepackage{longtable}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
%\usepackage{float}
\usepackage{url} % not crucial - just used below for the URL
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

A robust linear regression model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002) is used to treat the extreme values in the data. The robustness weight is used to determine if a value should be replaced with the fitted value from the model. This is constructed for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively. An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).

Figure \ref{fig:compare-plot} shows the profiles for a sample of 30 individuals the mean hourly wage before and after extreme values are replaced. The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.

The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers. It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time. To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed. Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined. Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals. This was achieved using a purpose written \texttt{shiny} (Chang et al. 2020) app (provided with the code of this paper). Observations where a change was made are tracked with a new variable called \texttt{is\_pred}, so that this effect in the downstream analyses can be monitored.

@@

\begin{itemize}
\item
  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://github.com/numbats/yowie/articles/process-data.html}. The code for the extreme value handling with robust linear fits is in \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html}.
\item
  \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed from \url{https://github.com/numbats/yowie}.
\item
  \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model. This app can be accessed at \url{https://ebsmonash.shinyapps.io/yowie_app/} with the source code provided \url{https://github.com/numbats/yowie/tree/master/app}.
\end{itemize}

@@
